---
title: "Getting started with Matisse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Matisse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

## What does Matisse do?

Genes can be spliced in different ways: certain exons may be included in some
transcripts but skipped in others. This is called **alternative splicing**, and
it means one gene can produce multiple protein isoforms with different functions.

Matisse measures alternative splicing **one cell at a time**. For each cell in
your dataset, it calculates a **PSI value** (Percent Spliced In) for each
splicing event you care about:

- **PSI = 1** — every transcript in that cell includes the exon
- **PSI = 0** — every transcript skips the exon
- **PSI = 0.5** — half include, half skip

By comparing PSI values across your cell types, clusters, or conditions, you can
ask which cell populations splice genes differently — and by how much.

Matisse slots directly into your existing
[Seurat](https://satijalab.org/seurat/) workflow. Your gene expression data,
UMAP, and cluster assignments are kept intact; splicing information is simply
added alongside them.

## Installation

```r
# install.packages("remotes")
remotes::install_github("avisrilab/Matisse")
```

## What data do I need?

Matisse supports two starting points depending on how your data were processed:

| My data | What I have | Path to use |
|---------|------------|-------------|
| 10x short-read (aligned with STAR) | Junction counts + Seurat object | `CreateMatisseObject()` |
| Long-read or Bagpiper transcript counts | Transcript counts + Seurat object | `CreateMatisseObjectFromTranscripts()` |

Both paths produce the same kind of Matisse object, and all downstream steps
(QC, filtering, visualization) are identical.

## Path 1: Short-read data (junction counts)

### Step 1 — Prepare your data

You need three things:

1. **A Seurat object** — your normal single-cell gene expression object, already
   processed (normalized, clustered, UMAP computed)
2. **A junction count table** — records how many reads span each exon-exon
   junction in each cell; produced by STARsolo alongside your gene counts
3. **A splice event table** — defines which junctions belong to which splicing
   event; generated by tools like rMATS, SUPPA2, or the built-in
   `BuildSimpleEvents()` helper

```{r eval=FALSE}
library(Matisse)
library(Seurat)
library(Matrix)

# --- Toy data for illustration ------------------------------------------------
set.seed(42)
n_cells <- 200; n_genes <- 500; n_jxns <- 50

# Simulate a gene expression matrix and process it with Seurat
gene_counts <- matrix(rpois(n_genes * n_cells, 10),
                      nrow     = n_genes,
                      dimnames = list(paste0("Gene", seq_len(n_genes)),
                                      paste0("CELL",  seq_len(n_cells))))

seu <- CreateSeuratObject(gene_counts) |>
  NormalizeData()        |>
  FindVariableFeatures() |>
  ScaleData()            |>
  RunPCA()               |>
  RunUMAP(dims = 1:10)   |>
  FindNeighbors(dims = 1:10) |>
  FindClusters()

# Simulate junction counts (one row per cell, one column per junction)
jxn_counts <- Matrix(
  matrix(rpois(n_cells * n_jxns, lambda = 3),
         nrow     = n_cells,
         dimnames = list(
           colnames(seu),
           paste0("chr1:", seq(1000, by = 500, length.out = n_jxns),
                  "-",     seq(1500, by = 500, length.out = n_jxns), ":+")
         )),
  sparse = TRUE
)
```

### Step 2 — Define splice events

A splice event says: *"for this exon, these junctions support inclusion and
these support skipping."* If you don't have a pre-built event table from rMATS
or SUPPA2, the `BuildSimpleEvents()` helper creates a simple one automatically
from your junction list.

```{r eval=FALSE}
# Junction annotation table (one row per junction)
jd <- data.frame(
  junction_id = colnames(jxn_counts),
  chr         = "chr1",
  start       = seq(1000, by = 500, length.out = n_jxns),
  end         = seq(1500, by = 500, length.out = n_jxns),
  strand      = "+",
  gene_id     = rep(paste0("Gene", 1:10), each = 5)
)

# Build events: each junction treated as one event vs. all others in the gene
ed <- BuildSimpleEvents(jd)
head(ed)
```

### Step 3 — Create a Matisse object

This combines your Seurat object, junction counts, and event definitions into
one place. Your gene expression data and cluster assignments are preserved
exactly as they were.

```{r eval=FALSE}
obj <- CreateMatisseObject(
  seurat          = seu,
  junction_counts = jxn_counts,
  event_data      = ed,
  junction_data   = jd
)
obj
```

### Step 4 — Calculate PSI

For each cell and each splicing event, Matisse counts the inclusion and
exclusion reads and computes the PSI ratio. Cells with fewer than
`min_coverage` total reads for an event are marked as missing (not enough
data to call a reliable splicing ratio).

```{r eval=FALSE}
obj <- CalculatePSI(obj, min_coverage = 5)
```

## Path 2: Transcript counts (Bagpiper / long-read)

If you have isoform-level counts (e.g. from Bagpiper with GENCODE transcript
IDs), provide your count table and SUPPA2 `.ioe` annotation files. Matisse
derives the PSI values from transcript abundance rather than junction reads.

```{r eval=FALSE}
# transcript_counts: rows = GENCODE transcript IDs, columns = cell barcodes
obj <- CreateMatisseObjectFromTranscripts(
  seurat            = seu,
  transcript_counts = transcript_counts,
  ioe_files         = c("events_SE.ioe", "events_RI.ioe"),
  min_coverage      = 5L
)
```

All steps below apply equally to both paths.

---

## Quality control

Before analysing splicing patterns, remove cells and events that don't have
enough data to be reliable.

### Calculate QC metrics

This adds per-cell summary statistics to your object: how many junctions were
detected, how many splicing events had enough reads, and the average PSI across
all covered events.

```{r eval=FALSE}
obj <- ComputeIsoformQC(obj)

# Visualise the QC metrics split by cluster
PlotQCMetrics(obj, group_by = "seurat_clusters")
```

### Remove low-quality cells

Keep only cells that have enough splicing data to be informative:

```{r eval=FALSE}
obj <- FilterCells(
  obj,
  min_junctions      = 5,   # at least 5 junctions detected
  min_junction_reads = 20,  # at least 20 total junction reads
  min_pct_covered    = 10   # PSI measurable for at least 10% of events
)
```

### Remove poorly covered events

Remove splicing events that were measurable in too few cells or that show no
variation across the dataset (everyone splices the same way — not interesting):

```{r eval=FALSE}
obj <- FilterEvents(
  obj,
  min_cells_covered = 20,    # event must be measurable in at least 20 cells
  min_psi_variance  = 0.01   # event must show some variation across cells
)
```

---

## Visualisation

### Overlay splicing on your UMAP

Colour each cell by its PSI value for a specific event. Cells with insufficient
coverage are shown in grey.

```{r eval=FALSE}
PlotPSIUMAP(obj, event_id = "Gene1:chr1:1000-1500:+:chr1:1000-1500:+")
```

### Compare splicing across cell types

```{r eval=FALSE}
PlotPSIViolin(
  obj,
  event_id = "Gene1:chr1:1000-1500:+:chr1:1000-1500:+",
  group_by = "seurat_clusters"
)
```

### Survey many events at once

```{r eval=FALSE}
# Heatmap of PSI values across cells and events
PlotPSIHeatmap(obj, max_cells = 100)
```

### Inspect junction usage for a gene

```{r eval=FALSE}
# Bar chart of total reads per junction for a gene of interest
PlotJunctionCoverage(obj, gene = "Gene1")
```

---

## Accessing your data

The Seurat object and all its metadata remain fully accessible at any time:

```{r eval=FALSE}
# Get the Seurat object back
seu_embedded <- GetSeurat(obj)

# Access cell metadata (e.g. cluster assignments) with $
obj$seurat_clusters

# Subset to specific cells — both gene expression and splicing data update together
subset_obj <- obj[c("CELL1", "CELL2", "CELL3"), ]
```

---

## Session info

```{r}
sessionInfo()
```
