---
title: "Getting started with Matisse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Matisse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What does Matisse do?

Genes can be spliced in different ways — certain exons may be included in some
transcripts but skipped in others. This is called **alternative splicing**, and
it means one gene can produce multiple protein variants with different
functions.

Matisse measures alternative splicing **one cell at a time**. For each cell it
calculates a **PSI value** (Percent Spliced In) for each splicing event:

- **PSI = 1** — every transcript in that cell includes the exon
- **PSI = 0** — every transcript skips the exon
- **PSI = 0.5** — half include, half skip

By comparing PSI values across cell types, clusters, or conditions you can
find which populations splice genes differently — and by how much. Matisse
sits on top of your existing [Seurat](https://satijalab.org/seurat/)
workflow. Your gene expression data, UMAP, and cluster labels stay intact;
splicing information is added alongside them.

---

## Two entry points

Matisse supports two data types that require slightly different starting steps:

| Data type | Entry point | When to use |
|-----------|-------------|-------------|
| **Short-read 10x Chromium** (STARsolo junction counts) | `CreateMatisseObject()` | You have a cells × junctions count matrix from STARsolo `--soloFeatures SJ` |
| **Long-read / transcript quantification** (Bagpiper, FLAMES, LIQA) | `CreateMatisseObjectFromTranscripts()` | You have a transcripts × cells count matrix and SUPPA2 IOE files |

Once the object is built, all downstream functions — QC, filtering,
normalisation, visualisation — are identical for both data types.

---

## Quick-start (short reads)

```{r short-read-quick, eval=FALSE}
library(Matisse)

obj <- CreateMatisseObject(
  seurat          = seu,          # your existing Seurat object
  junction_counts = jxn_counts,   # cells × junctions from STARsolo
  event_data      = event_df      # splice event annotation
)
obj <- CalculatePSI(obj, min_coverage = 5)
obj <- ComputeIsoformQC(obj)
PlotPSIUMAP(obj, event_id = "PTBP1:SE:chr18:3433647-3436055")
```

→ **Full walkthrough:** [Short-read workflow](short-reads.html)

---

## Quick-start (long reads)

```{r long-read-quick, eval=FALSE}
library(Matisse)

obj <- CreateMatisseObjectFromTranscripts(
  seurat            = seu,
  transcript_counts = tx_counts,    # transcripts × cells from Bagpiper/FLAMES
  ioe_files         = "events_SE.ioe",
  min_coverage      = 5L
)
# PSI is already computed; normalise transcripts then cluster
obj <- SCTransformTranscripts(obj, n_pca_dims = 50)
obj <- RunUMAP(obj, dims = 1:50)
obj <- FindNeighbors(obj, dims = 1:50)
obj <- FindClusters(obj, resolution = 0.5)
PlotPSIUMAP(obj, event_id = "SE:chr18:3433647-3436055:+")
```

→ **Full walkthrough:** [Long-read workflow](long-reads.html)

---

## Session info

```{r session-info}
sessionInfo()
```
