---
title: "Introduction to Matisse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Matisse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

## Overview

**Matisse** (Multi-modal Analysis of Transcript Isoforms in Single-cell
Sequencing Experiments) provides an integrated framework for analysing
isoform-resolved single-cell RNA-seq data. It builds on
[Seurat](https://satijalab.org/seurat/) and
[Signac](https://stuartlab.org/signac/), adding:

- A `MatisseObject` that wraps a `Seurat` object and co-stores isoform layers
- PSI (Percent Spliced In) matrix calculation from junction read counts
- Isoform-aware quality control metrics and filtering
- Visualization functions for PSI values on UMAPs, violins, and heatmaps

## Installation

```r
# Install development version from GitHub
# install.packages("remotes")
remotes::install_github("avisrilab/Matisse")
```

## The MatisseObject

The central data structure is a `MatisseObject`. It consists of:

| Slot | Contents |
|------|----------|
| `seurat` | An embedded `Seurat` object (gene expression, reductions, metadata) |
| `junction_counts` | Sparse matrix: cells × junctions |
| `psi` | Sparse matrix: cells × splice events (PSI in [0, 1]) |
| `inclusion_counts` | Sparse matrix: cells × events |
| `exclusion_counts` | Sparse matrix: cells × events |
| `event_data` | Splice event annotation table |
| `junction_data` | Junction annotation table |
| `isoform_metadata` | Per-cell isoform QC metrics |

## Quick Start

### 1. Prepare your data

Matisse expects:

1. A `Seurat` object (processed gene-level data from any pipeline)
2. A sparse junction count matrix (cells × junctions), e.g. from
   STARsolo, Cell Ranger, or a long-read quantifier
3. A splice event annotation table (e.g. exported from rMATS or SUPPA2)

```{r eval=FALSE}
library(Matisse)
library(Seurat)
library(Matrix)

# -- Toy data for illustration -----------------------------------------------
set.seed(42)
n_cells <- 200; n_genes <- 500; n_jxns <- 50

gene_counts <- matrix(rpois(n_genes * n_cells, 10),
                      nrow     = n_genes,
                      dimnames = list(paste0("Gene", seq_len(n_genes)),
                                      paste0("CELL", seq_len(n_cells))))

seu <- CreateSeuratObject(gene_counts) |>
  NormalizeData() |>
  FindVariableFeatures() |>
  ScaleData() |>
  RunPCA() |>
  RunUMAP(dims = 1:10) |>
  FindNeighbors(dims = 1:10) |>
  FindClusters()

# Junction counts (cells × junctions)
jxn_counts <- Matrix(
  matrix(rpois(n_cells * n_jxns, lambda = 3),
         nrow     = n_cells,
         dimnames = list(colnames(seu),
                         paste0("chr1:", seq(1000, by = 500, length.out = n_jxns),
                                "-", seq(1500, by = 500, length.out = n_jxns), ":+"))),
  sparse = TRUE
)
```

### 2. Define splice events

```{r eval=FALSE}
# Build a simple one-versus-rest event table from the junction list
jd <- data.frame(
  junction_id = colnames(jxn_counts),
  chr         = "chr1",
  start       = seq(1000, by = 500, length.out = n_jxns),
  end         = seq(1500, by = 500, length.out = n_jxns),
  strand      = "+",
  gene_id     = rep(paste0("Gene", 1:10), each = 5)
)

# Use the built-in helper for simple (one-vs-rest) events
ed <- BuildSimpleEvents(jd)
head(ed)
```

### 3. Create a MatisseObject

```{r eval=FALSE}
obj <- CreateMatisseObject(
  seurat          = seu,
  junction_counts = jxn_counts,
  event_data      = ed,
  junction_data   = jd
)
obj
```

### 4. Calculate PSI

```{r eval=FALSE}
obj <- CalculatePSI(obj, min_coverage = 5)
```

### 5. Quality control

```{r eval=FALSE}
obj <- ComputeIsoformQC(obj)
PlotQCMetrics(obj, group_by = "seurat_clusters")
```

Filter out cells with insufficient isoform coverage:

```{r eval=FALSE}
obj <- FilterCells(
  obj,
  min_junctions      = 5,
  min_junction_reads = 20,
  min_pct_covered    = 10
)
```

Remove poorly covered events:

```{r eval=FALSE}
obj <- FilterEvents(obj, min_cells_covered = 20, min_psi_variance = 0.01)
```

### 6. Visualisation

```{r eval=FALSE}
# UMAP coloured by PSI of a specific event
PlotPSIUMAP(obj, event_id = "Gene1:chr1:1000-1500:+:chr1:1000-1500:+")

# Violin plot across clusters
PlotPSIViolin(obj,
              event_id = "Gene1:chr1:1000-1500:+:chr1:1000-1500:+",
              group_by = "seurat_clusters")

# Heatmap of top variable events
PlotPSIHeatmap(obj, max_cells = 100)

# Per-gene junction coverage
PlotJunctionCoverage(obj, gene = "Gene1")
```

## Accessing the embedded Seurat object

All Seurat functionality remains directly accessible:

```{r eval=FALSE}
# Extract the Seurat object
seu_embedded <- GetSeurat(obj)

# Seurat-style metadata access via $
obj$seurat_clusters

# Subset by cell barcode (keeps both Seurat and isoform layers in sync)
subset_obj <- obj[c("CELL1", "CELL2", "CELL3"), ]
```

## Next steps

- **Vignette: PSI analysis** — differential splicing between cell types
- **Vignette: Integration with Signac** — linking splicing to chromatin
  accessibility in multiome data

## Session info

```{r}
sessionInfo()
```
