---
title: "Getting started with Matisse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Matisse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What does Matisse do?

Genes can be spliced in different ways -- certain exons may be included in some
transcripts but skipped in others. This is called **alternative splicing**, and
it means one gene can produce multiple protein variants with different
functions.

Matisse measures alternative splicing **one cell at a time**. For each cell it
calculates a **PSI value** (Percent Spliced In) for each splicing event:

- **PSI = 1** -- every transcript in that cell includes the exon
- **PSI = 0** -- every transcript skips the exon
- **PSI = 0.5** -- half include, half skip

By comparing PSI values across cell types, clusters, or conditions you can
find which populations splice genes differently -- and by how much. Matisse
sits on top of your existing [Seurat](https://satijalab.org/seurat/)
workflow. Your gene expression data, UMAP, and cluster labels stay intact;
splicing information is added alongside them.

---

## Two entry points

Matisse supports two data types via a single constructor. What you pass in
determines the mode:

| Data type | How to create | When to use |
|-----------|--------------|-------------|
| **Long-read / transcript quantification** (Bagpiper, FLAMES, LIQA) | `CreateMatisseObject(transcript_counts=..., events=...)` | You have a transcripts x cells count matrix and SUPPA2 IOE files |
| **Short-read 10x Chromium** (STARsolo junction counts) | `CreateMatisseObject(junction_counts=..., events=...)` | You have a cells x junctions count matrix from STARsolo `--soloFeatures SJ` |

Once the object is built, all downstream functions -- QC, filtering,
normalisation, visualisation -- are identical for both data types.

---

## Quick-start (long reads)

```{r long-read-quick, eval=FALSE}
library(Matisse)

# transcript_counts: transcripts x cells (e.g. from Bagpiper, FLAMES, LIQA)
# events: SUPPA2 .ioe file path(s); CreateMatisseObject parses internally.
# PSI is computed automatically at construction.
obj <- CreateMatisseObject(
  seurat            = seu,
  transcript_counts = tx_counts,
  events            = "events_SE.ioe",
  min_coverage      = 5L
)

# Summarise PSI per event, then normalise + reduce + cluster on transcripts
psi_summary <- SummarizePSI(obj)
obj <- SCTransform(obj)
obj <- RunPCA(obj, assay = "SCT", npcs = 50)
obj <- RunUMAP(obj, dims = 1:50)
obj <- FindNeighbors(obj, dims = 1:50)
obj <- FindClusters(obj, resolution = 0.5)
PlotUMAP(obj, feature = "SE:chr18:3433647-3436055:+")
```

-> **Full walkthrough:** [Long-read workflow](long-reads.html)

---

## Quick-start (short reads)

```{r short-read-quick, eval=FALSE}
library(Matisse)

# junction_counts: cells x junctions matrix from STARsolo (--soloFeatures SJ).
# Junction IDs encode coordinates (e.g. "chr1-12345-67890-+"), auto-parsed
# for sashimi plots.
# events: data.frame OR path(s) to a SUPPA2-style .ioe file.
obj <- CreateMatisseObject(
  seurat          = seu,
  junction_counts = jxn_counts,
  events          = event_df,
  min_coverage    = 5L
)
PlotUMAP(obj, feature = "PTBP1:SE:chr18:3433647-3436055")
```

-> **Full walkthrough:** [Short-read workflow](short-reads.html)

---

## Session info

```{r session-info}
sessionInfo()
```
