---
title: "Long-read workflow: transcript-based PSI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Long-read workflow: transcript-based PSI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## When to use this workflow

Use this workflow when your data was generated with a long-read platform
(PacBio, Oxford Nanopore) or when you have used a short-read isoform
quantifier such as **Bagpiper**, **FLAMES**, or **LIQA** that produces
per-transcript per-cell counts. These tools assign each read to a specific
full-length transcript rather than counting at the junction level.

If you have STARsolo junction counts from standard 10x Chromium short reads,
see the [Short-read workflow](short-reads.html) instead.

---

## What you need

| Input | Description |
|-------|-------------|
| **Seurat object** | Already processed: gene-expression normalisation, UMAP, cluster labels. |
| **Transcript count matrix** | *Transcripts × cells* matrix of UMI counts. Row names are transcript IDs matching those in your annotation. |
| **SUPPA2 IOE files** | One or more `.ioe` files from SUPPA2's `generateEvents` command, one per event type (SE, RI, SS, MX, FL). These map transcripts to the splicing events they support. |

---

## Step 1 -- Build the Matisse object (PSI is computed in the same call)

Pass `transcript_counts` and `events` together. The `events` parameter
accepts either a character vector of SUPPA2 `.ioe` file paths (parsed
internally) or a pre-built `data.frame`. Matisse stores the transcript
counts in an internal assay and computes PSI as part of construction.

```{r build-object, eval=FALSE}
library(Matisse)

# transcript_counts: transcripts x cells sparse matrix (e.g. from Bagpiper)
# events: SUPPA2 .ioe file paths, one per event type
obj <- CreateMatisseObject(
  seurat            = seu,
  transcript_counts = transcript_counts,
  events            = c(
    "events_SE.ioe",   # skipped exons
    "events_RI.ioe",   # retained introns
    "events_SS.ioe"    # alternative splice sites
  ),
  min_coverage      = 5L
)
```

For each cell and event, Matisse sums transcript counts from inclusion
isoforms and exclusion isoforms, then computes:

$$PSI_{c,e} = \frac{\sum \text{inclusion transcript counts}}
                   {\sum \text{inclusion counts} + \sum \text{exclusion counts}}$$

Cells with fewer than `min_coverage` total transcripts for a given event
are reported as `NA`. After construction, three QC columns sit in cell
metadata:

- `nCount_isoform` -- total transcript counts per cell
- `nFeature_isoform` -- number of transcripts with ≥1 read per cell
- `nPercent_isoform` -- percentage of splice events with a non-NA PSI per cell

To re-compute PSI with different parameters later, call `CalculatePSI(obj,
min_coverage = ...)`. To skip the PSI step at construction (rare), pass
`defer_psi = TRUE` and call `CalculatePSI()` manually.

---

## Step 2 -- Quality control

### Visualise QC metrics

Call `PlotViolin()` with no `feature` argument to automatically plot all
three isoform QC metrics (`nCount_isoform`, `nFeature_isoform`,
`nPercent_isoform`) as a faceted panel. This is the fastest way to
spot cells with very few transcripts or low splicing coverage.

```{r qc-plot, eval=FALSE}
PlotViolin(obj)
```

### Remove low-quality cells

Remove cells that have too few transcripts or too little splicing coverage:

```{r filter-cells, eval=FALSE}
obj <- FilterCells(
  obj,
  min_features_isoform = 5,    # at least 5 transcripts detected
  min_pct_isoform      = 10    # PSI covered for >= 10% of splice events
)
```

### Remove uninformative splice events

Remove events that are observed in very few cells:

```{r filter-events, eval=FALSE}
obj <- FilterEvents(obj, min_cells_covered = 20)
```

---

## Step 3 -- Normalise transcript counts

Normalise the transcript-level count matrix before clustering. SCTransform
applies variance stabilisation and corrects for sequencing depth.

```{r sctransform, eval=FALSE, downlit=FALSE}
obj <- SCTransform(obj)
obj <- RunPCA(obj, assay = "SCT", npcs = 50)
```

---

## Step 4 -- Cluster and embed

Standard Seurat clustering functions work directly on the Matisse object.
The PSI data and all other slots are preserved throughout.

```{r cluster, eval=FALSE, downlit=FALSE}
obj <- RunUMAP(obj, dims = 1:50)
obj <- FindNeighbors(obj, dims = 1:50)
obj <- FindClusters(obj, resolution = 0.5)
```

---

## Step 5 -- Visualise splicing patterns

### Overlay PSI on the UMAP

Each dot is a cell coloured by its PSI value for one splice event:
blue = exon skipped (low PSI), red = exon included (high PSI).

```{r umap, eval=FALSE}
PlotUMAP(
  obj,
  feature = "SE:chr18:3433647-3436055:+",
  title   = "PTBP1 exon 9 -- PSI per cell"
)
```

### Compare splicing between cell types

```{r violin, eval=FALSE}
PlotViolin(
  obj,
  feature  = "SE:chr18:3433647-3436055:+",
  group_by = "cell_type"
)
```

### Survey all events at once

```{r heatmap, eval=FALSE}
PlotHeatmap(obj, group_by = "seurat_clusters", max_cells = 400)
```

### Inspect isoform usage with a sashimi plot

`PlotSashimi()` draws junction arcs scaled by aggregate read count,
coloured by role (inclusion = blue, exclusion = red). Facet by cell type
to compare isoform usage across populations.

```{r coverage, eval=FALSE}
PlotSashimi(
  obj,
  event_id = "SE:chr18:3433648-3434699:3434801-3436055:-",
  group_by = "cell_type",
  title    = "PTBP1 exon 9 -- coverage by cell type"
)
```

> **Note on event IDs:** SUPPA2 SE event IDs use the format
> `SE:chr:donor1-acceptor1:donor2-acceptor2:strand`
> (e.g. `SE:chr18:3433648-3434699:3434801-3436055:-`). These become
> the feature names in `GetPSI(obj)`, so use the same string when calling
> `PlotUMAP()`, `PlotSashimi()`, or subsetting.

---

## Accessing data at any point

```{r access, eval=FALSE}
# Extract the underlying Seurat object
seu <- GetSeurat(obj)

# Cell metadata via $ (forwarded from Seurat)
obj$seurat_clusters
obj$cell_type

# Full PSI matrix as a sparse matrix (cells x splice events)
psi <- GetPSI(obj)

# Raw transcript counts (transcripts x cells) via the native Seurat layer API
tx <- SeuratObject::GetAssayData(GetSeurat(obj)[["isoform"]], "counts")

# Subset to one cell type -- PSI and gene expression stay in sync
neurons <- obj[obj$cell_type == "Neuron", ]
```

---

## Session info

```{r session-info}
sessionInfo()
```
